Unshuffling Permutations

### 2015-09-25 submission.tex
Conclusion

      \draw[step=1cm,black!30,thin,fill=black!10] (0,0) grid (12,12);

@@

\end{proof}

There are a number of further directions of investigation in this
general subject. We mention several -~not necessarily new~- open
problems that are, in our opinion, the most interesting. How many
permutations of $S_{2n}$ are squares? How many $(213,231)$-avoiding
permutations of $S_{2n}$ are squares? (Equivalently, by
Proposition~\ref{prop:bijection_binary_to_permutations_squares},
how many binary strings of length $2n$ are squares; Problem~4
in \cite{Henshall:Rampersad:Shallit:2011})? How hard is the problem of
deciding whether a $(213,231)$-avoiding permutation is a square
(Problem~4 in \cite{Henshall:Rampersad:Shallit:2011},
see also \cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013})?
Given two permutations $\pi$ and~$\sigma$, how hard is the problem of
deciding whether $\sigma$ is a square root of~$\pi$?

Besides, in this work, we investigate some algebraic properties of the
shuffle of permutations which have direct combinatorial consequences.
However, this work raises several additional classical algebraic
questions. Among these, one can ask for a complete algebraic study of
$\QQ[S]$ as a graded associative algebra for the shuffle  product
$\SHUFFLE$. Describing a generating family for $\QQ[S]$, defining
multiplicative bases of $\QQ[S]$, and determining whether $\QQ[S]$ is
free as an associative algebra are worthwhile questions.

  As for \emph{2}, suppose, still aiming at a contradiction, that there
  exist two $(\sigma_1, \sigma_2)$-edges $(a, a')$ and $(b, b')$ in
  $\mathcal{M}$. Since $\mathcal{M}$ is containment-free, we may assume
  $a < b$ and $a' < b'$. Therefore, we have $\pi(a) > \pi(b)$ (since
  $\sigma_1$ is decreasing) and $\pi(a') < \pi(b')$ (since $\sigma_2$ is
  increasing), which contradicts our hypothesis about~$\mathcal{M}$.
  \qed
\end{proof}

\bigskip

\begin{proof}[of Proposition~\ref{proposition:hardness}]
  Suppose first that $\sigma$ occurs in $\pi$ and fix an occurrence.
  Construct a matching $\mathcal{M}$ as follows:
  \begin{itemize}
    \item $\mathcal{M}$ contains $N_1$ pairwise crossing
    $(\nu_1, \nu'_1)$-edges.
    \item $\mathcal{M}$ contains $N_2$ pairwise crossing
    $(\nu_2, \nu'_2)$-edges.
    \item $\mathcal{M}$ contains $N_3$ pairwise crossing
    $(\nu_3, \nu'_3)$-edges.
    \item $\mathcal{M}$ contains $N_4$ pairwise crossing
    $(\nu_4, \nu'_4)$-edges.
    \item $\mathcal{M}$ contains $k+2$ pairwise crossing
    $(\sigma', \pi')$-edges as depicted in
    Figure~\ref{fig:subfig:sigma' - pi' - pi'' - sigma''}.
    More precisely,
    (i) the first position of $\sigma'$
    (\emph{i.e.}, $(2N_1+N_2+N_3) + 1$) is linked
    to the first position of $\pi'$
    (\emph{i.e.}, $(2N_1 + 2N_2 + 2N_3 + N_4 + k + 2) + 1$),
    (ii) the last position of $\sigma'$
    (\emph{i.e.}, $(2N_1+N_2+N_3) + k+2$) is linked
    to the last position of $\pi'$
    (\emph{i.e.}, $(2N_1 + 2N_2 + 2N_3 + N_4 + k + 2) + n+2$),
    and all other positions in $\sigma'$ are linked by means of $k$ pairwise
    crossing edges to the positions in
    $\pi'$ that correspond to the fixed occurence of $\sigma$ in $\pi$.
    (Notice that we use here the fact that $\sigma$ occurs in $\pi$).
    \item $\mathcal{M}$ contains $n-k$ pairwise crossing
    $(\pi', \pi'')$-edges as depicted in
    Figure~\ref{fig:subfig:sigma' - pi' - pi'' - sigma''}.
    More precisely,
    all positions in $\pi'$ that do not correspond to the fixed occurence of
    $\sigma$ in $\pi$ are linked by means of $n-k$ pairwise crossing edges
    to the positions in $\pi''$ that do not correspond to the fixed
    occurrence of $\sigma$ in $\pi$.
    \item $\mathcal{M}$ contains $k$ pairwise crossing
    $(\pi'', \sigma'')$-edges as depicted in
    Figure~\ref{fig:subfig:sigma' - pi' - pi'' - sigma''}.
    More precisely, the positions in $\pi''$ that correspond to
    the fixed occurrence of $\sigma$ in $\pi$ are llinked
    linked by means of $k$ pairwise crossing edges to all positions in
    $\sigma''$.
    (Notice that we use here the fact that $\sigma$ occurs in $\pi$).
  \end{itemize}
  It can be easily checked (probably referring to Figure~\ref{fig:reduction}) that
  $\mathcal{M}$ is perfect, containment-free and that
  for any two disctinct edges
  $(a, b)$ and $(a', b')$ in $\mathcal{M}$
  we have $\mu(a) < \mu(a')$ if and only if $\mu(b) < \mu(b')$.

@@

>>>>

      \draw[step=1cm,black!30,ultra thin,fill=black!10] (0,0) grid (12,12);

@@

  Suppose first that $\sigma$ occurs in $\pi$ and fix an occurrence.
  Construct a matching $\mathcal{M}$ as follows:
  \begin{itemize}
    \item $\mathcal{M}$ contains $N_1$ pairwise crossing
    $(\nu_1, \nu'_1)$-edges.
    \item $\mathcal{M}$ contains $N_2$ pairwise crossing
    $(\nu_2, \nu'_2)$-edges.
    \item $\mathcal{M}$ contains $N_3$ pairwise crossing
    $(\nu_3, \nu'_3)$-edges.
    \item $\mathcal{M}$ contains $N_4$ pairwise crossing
    $(\nu_4, \nu'_4)$-edges.
    \item $\mathcal{M}$ contains $k+2$ pairwise crossing
    $(\sigma', \pi')$-edges as depicted in
    Figure~\ref{fig:subfig:sigma' - pi' - pi'' - sigma''}.
    More precisely,
    (i) the first position of $\sigma'$
    (\emph{i.e.}, $(2N_1+N_2+N_3) + 1$) is linked
    to the first position of $\pi'$
    (\emph{i.e.}, $(2N_1 + 2N_2 + 2N_3 + N_4 + k + 2) + 1$),
    (ii) the last position of $\sigma'$
    (\emph{i.e.}, $(2N_1+N_2+N_3) + k+2$) is linked
    to the last position of $\pi'$
    (\emph{i.e.}, $(2N_1 + 2N_2 + 2N_3 + N_4 + k + 2) + n+2$),
    and all other positions in $\sigma'$ are linked by means of $k$ pairwise
    crossing edges to the positions in
    $\pi'$ that correspond to the fixed occurence of $\sigma$ in $\pi$.
    (Notice that we use here the fact that $\sigma$ occurs in $\pi$).
    \item $\mathcal{M}$ contains $n-k$ pairwise crossing
    $(\pi', \pi'')$-edges as depicted in
    Figure~\ref{fig:subfig:sigma' - pi' - pi'' - sigma''}.
    More precisely,
    all positions in $\pi'$ that do not correspond to the fixed occurence of
    $\sigma$ in $\pi$ are linked by means of $n-k$ pairwise crossing edges
    to the positions in $\pi''$ that do not correspond to the fixed
    occurrence of $\sigma$ in $\pi$.
    \item $\mathcal{M}$ contains $k$ pairwise crossing
    $(\pi'', \sigma'')$-edges as depicted in
    Figure~\ref{fig:subfig:sigma' - pi' - pi'' - sigma''}.
    More precisely, the positions in $\pi''$ that correspond to
    the fixed occurrence of $\sigma$ in $\pi$ are llinked
    linked by means of $k$ pairwise crossing edges to all positions in
    $\sigma''$.
    (Notice that we use here the fact that $\sigma$ occurs in $\pi$).
  \end{itemize}
  It can be easily checked (probably referring to Figure~\ref{fig:reduction}) that
  $\mathcal{M}$ is perfect, containment-free and that
  for any two disctinct edges
  $(a, b)$ and $(a', b')$ in $\mathcal{M}$
  we have $\mu(a) < \mu(a')$ if and only if $\mu(b) < \mu(b')$.

@@

% Unifier (une seule partie)

There are a number of further directions of investigation in this
general subject. They cover several areas: algorithmic, combinatorics,
and algebra. Let us mention several -~not necessarily new~- open
problems that are, in our opinion, the most interesting. How many
permutations of $S_{2n}$ are squares? How many $(213,231)$-avoiding
permutations of $S_{2n}$ are squares? (Equivalently, by
Proposition~\ref{prop:bijection_binary_to_permutations_squares},
how many binary strings of length $2n$ are squares; Problem~4
in \cite{Henshall:Rampersad:Shallit:2011})? How hard is the problem of
deciding whether a $(213,231)$-avoiding permutation is a square
(Problem~4 in \cite{Henshall:Rampersad:Shallit:2011},
see also \cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013})?
Given two permutations $\pi$ and~$\sigma$, how hard is the problem of
deciding whether $\sigma$ is a square root of~$\pi$?
%Besides, in this work, we investigate some algebraic properties of the
%shuffle of permutations which have direct combinatorial consequences.
%However, this work raises several additional classical algebraic
%questions. Among these,
Moreover, one can ask for a complete algebraic study of
$\QQ[S]$ as a graded associative algebra for the shuffle  product
$\SHUFFLE$. Describing a generating family for $\QQ[S]$, defining
multiplicative bases of $\QQ[S]$, and determining whether $\QQ[S]$ is
free as an associative algebra are worthwhile questions.

  As for \emph{2}, suppose, still aiming at a contradiction, that there
  exist two $(\sigma_1, \sigma_2)$-edges $(a, a')$ and $(b, b')$ in
  $\mathcal{M}$. Since $\mathcal{M}$ is containment-free, we may assume
  $a < b$ and $a' < b'$. Therefore, we have $\pi(a) > \pi(b)$ (since
  $\sigma_1$ is decreasing) and $\pi(a') < \pi(b')$ (since $\sigma_2$ is
  increasing), which contradicts our hypothesis about~$\mathcal{M}$.
  \qed
\end{proof}
\bigskip
